Coalescent inference of mutation spectrum histories from sample frequency spectra

### 2019-09-23 manuscript/main.tex
added various losses to manuscript

@@

%WD more inverse problem intro
The inverse problem \eqref{eqn:transform_discrete} is ill-posed in general, so many very different histories can be equally consistent with the data \citep{oscillation paper? Yun's other papers?}.
In order to recover smooth histories,
we penalize the $L^p$ norms
of the time derivative of $\boldsymbol\mu(t)$
\[
\sum_{i=1}^{\kappa_k}\left\| \frac{d \mu_i(t)}{d t} \right\|_p^p
= \sum_{i=1}^{\kappa_k}\int_0^\infty\left|\frac{d\mu_i(t)}{dt}\right|^p dt,
\]
for $p=1,2$.
With piecewise constant histories as in \eqref{eqn:transform_discrete} this becomes
\[
\left\|\Delta Z \right\|_p^p.
\]
where $\Delta$ denotes the first difference matrix.
Our cost function (penalized log-likelihood) is then
\begin{equation}
\label{eqn:penalized}
C(Z)
%\tilde\ell(\boldsymbol z)
= -\ell(Z) + \lambda \left(\alpha\left\|\Delta Z\right\|_1 + \frac{(1-\alpha)}{2}\left\|\Delta Z\right\|_2^2\right),
\end{equation}
where $\lambda$ tunes the regularization strength, and $\alpha$ tunes the relative $p=1$ vs $p=2$ regularization.
\textcolor{blue}{Kam note: probably we need a ``Ridge'' (Tikhonov)
term like the following to deal with small singular values:}
\[
\lambda' \| Z \|_2^2
\]

>>>>

Humblebrag: the word \emph{spectrum} is triply overloaded in this paper:
\begin{enumerate}
\item the traditional sample frequency \emph{spectrum}
\item mutation \emph{spectrum} as in the nucleotide context of each mutation
\item \emph{spectral} regularization for encouraging low rank solutions
\end{enumerate}

@@

%WD more inverse problem intro
The inverse problem \eqref{eqn:transform_discrete} is ill-posed in general, 
so many very different histories can be equally consistent with the data 
\citep{oscillation paper? Yun's other papers?}.
We deal with this problem using regularization, to enforce well-behaved histories.
Our cost function to minimize, a penalized log-likelihood, is
\begin{equation}
\label{eqn:penalized}
C(Z)
%\tilde\ell(\boldsymbol z)
= -\ell(Z) + R(Z) .
\end{equation}
The function $R(Z)$ incorporates our regularization.
In order to recover smooth histories,
we penalize the $L^p$ norms
of the time derivative of $\boldsymbol\mu(t)$
\[
\sum_{i=1}^{\kappa_k}\left\| \frac{d \mu_i(t)}{d t} \right\|_p^p
= \sum_{i=1}^{\kappa_k}\int_0^\infty\left|\frac{d\mu_i(t)}{dt}\right|^p dt,
\]
for $p=1,2$.
With piecewise constant histories as in \eqref{eqn:transform_discrete} this becomes
\[
\left\|\Delta Z \right\|_p^p.
\]
where $\Delta$ denotes the first difference matrix.
We have implemented three different types of regularization.

The first option we call ``soft'' rank regularization, with 
\begin{equation}
  \label{eq:regularization_soft}
  R_\mathrm{soft}(Z) 
  = 
  \lambda_\mathrm{spline} 
  \frac{1}{2} \left\|\Delta Z\right\|_2^2
  +
  \lambda_\mathrm{rank} \| Z \|_*.
\end{equation}
The spline term controls the smoothness of solutions, 
and the rank term allows us to recover low-rank solutions.
The rank penalty is the nuclear norm $\| Z \|_*$, 
which is equivalent to the $\ell^1$-norm on the singular values.
If $Z = U S V^\intercal$ is the SVD of $Z$, with $S = \mathrm{diag}(\sigma)$ then $\| Z \|_* = \| \sigma \|_1$.
The parameters $\lambda_\mathrm{spline}$ and $\lambda_\mathrm{rank}$ control the overall strength
of smoothing and rank regularization.
The function $R_\mathrm{soft}$ is convex; therefore we have a convex optimization problem 
in this scenario.

The second option is ``hard'' rank regularization, where
\begin{equation}
  \label{eq:regularization_soft}
  R_\mathrm{hard}(Z) 
  = 
  \lambda_\mathrm{spline} 
  \frac{1}{2} \left\|\Delta Z\right\|_2^2
  +
  \lambda_\mathrm{rank} \, \mathrm{rank}(Z).
\end{equation}
Here, the rank penalty is the actual rank of $Z$,
and note that $\mathrm{rank}(Z) = \| \sigma \|_0$.
In this case the problem is {\em not} convex.
However, we may still compute a proximal map for the non-differentiable rank term
and optimize with prox-gradients.

The third option allows more general smoothing but ignores all rank penalization:
\begin{equation}
  \label{eq:regularization_soft}
  R_\mathrm{indep}(Z) 
  = 
  \lambda_\mathrm{TV} 
  \left(
    \alpha_\mathrm{TV}\left\|\Delta Z\right\|_1 + \frac{(1-\alpha_\mathrm{TV})}{2}\left\|\Delta Z\right\|_2^2
  \right) .
\end{equation}
The TV terms control the smoothness of solutions, 
and the rank term allows us to recover low-rank solutions.
The function $R_\mathrm{indep}$ is convex; 
therefore we have a convex optimization problem.
Note also that the columns of $Z$ are uncoupled, 
so that we may solve this case as $\kappa_k$ different subproblems.
